---
title: "Reproducible script for the publication"
author: "Weinholdt Claus"
date: "2018-24-4"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Weinholdt et. al __Analysis of genes regulated by isoforms of the epidermal growth factor receptor in a glioblastoma cell line__

## Contents

* [Getting started](#start)
    * [Installation](#Installation)
    * [Data](#data)
* [Visualization](#vis)

## <a name="start"></a>Getting started

### <a name="Installation"></a>Installation of necessary packages

Install and load packages containing the functions for the analyses

```r
install.packages("devtools")
devtools::install_github("BGSC")
library(BGSC)
```

### <a name="data"></a> Load expression data
Loading CXXX data 
```r
data(ExpData)
```

```{r, comment=NA, echo=FALSE}
suppressWarnings( library("BGSC",quietly = TRUE,warn.conflicts = FALSE) )
library(ggplot2)
library(gridExtra)
library(grid)
library(gtable)
```

### normalize Data
```{r normalizeExpData}
normData <- normalizeExpData()
```

### calulate logLik for class  Data
```{r logLik}
Lsets <- get.Lset()
normDataLogLik <- logLikelihoodOfnormData(normData$E)
rownames(ALL.MUs) <- rownames(normData$E)
colnames(ALL.MUs) <- c('a0','a1','b0','b1','c0','c1','d0','d1')
rownames(ALL.VARs) <- rownames(normData$E)
colnames(ALL.VARs) <- c('a','b','c','d')
```

#### Density plots with fitted normal distribution 

```{r Density, echo=FALSE, fig.height=9, fig.width=9}
aaa <- Density.NV.fit.plot(id = 'ILMN_1687840',normData,useGroup = "a",DOplot = FALSE)
bbb <- Density.NV.fit.plot(id = 'ILMN_1684585',normData,useGroup = "b",DOplot = FALSE)
ccc <- Density.NV.fit.plot(id = 'ILMN_1730999',normData,useGroup = "c",DOplot = FALSE)
ddd <- Density.NV.fit.plot(id = 'ILMN_2320964',normData,useGroup = "d",DOplot = FALSE)

grid.arrange(aaa + theme(legend.position="none")  , 
             bbb + theme(legend.position="none") ,
             ccc + theme(legend.position="none") ,
             ddd + theme(legend.position="none") ,
             ncol=2,nrow=2)
```

### calulate BIC from logLik 
```{r BIC}
  npar <- sapply(Lsets, function(x) sum(!sapply(x,is.null ) )) + 1  ## number parapeters for LogLilk -> mean + var 
  k <-  sapply(Lsets, function(x) sum( sapply(x,length) ))
  normDataBIC <- get.IC(normDataLogLik , npar, k , IC = 'BIC')
  BICminInd <- apply( normDataBIC,MARGIN = 1, FUN = minIndex)
  print(table(BICminInd))
``` 

###  calulate Posterior from BIC
```{r Posterior}
  normDataPosterior <- get.Posterior( normDataBIC ,Pis = c(0.7,0.1,0.1,0.1))
  POSTmaxInd <- apply(normDataPosterior, MARGIN = 1 ,FUN = maxIndex)
  print(table(POSTmaxInd))
```

###  get gene classes
```{r PostClass}   
  PostClass <- get.gene.classes(data = normDataPosterior,indexing = "max",filter = 0.75, DoPlot = TRUE)
```   

## Comapre Illumina data with RT-qPCR
```{r load qPCR}
 ### compare to qPCR ----------------------------------------------------------------------
    MeanFoldChangeClass <- get.log2Mean.and.log2FC(normData = normData)
    qCPRdataC <- getQPCR()
    
    IDs.dt <- data.table::data.table(normData$genes,keep.rownames = T,key = 'rn')
    IDs.dt.c <- IDs.dt[rownames(PostClass$resFilter$c),]
    data.table::setkey(IDs.dt.c,'SYMBOL')
    
    qgenesIDs <- lapply(qCPRdataC$rn, function(qg) as.character(IDs.dt.c[qg,][['rn']]) )
    names(qgenesIDs) <- qCPRdataC$rn
  
    print(do.call(c,qgenesIDs))
```

```{r barplots data ,echo=FALSE}        

    make.plot.data.FC.Ill.qPCR <- function(qCPRdata,MeanFoldChangeClass, class="c"){
      TMP <- data.frame()
      for (tmpG in qCPRdataC$rn  ) {
        expC <- MeanFoldChangeClass[[class]][qgenesIDs[[tmpG]], ]
        TMPexp <- data.frame( "Gene" = tmpG,
                              "ExpID" = expC$rn,
                              "Set"="Illumina",
                              "FC" =expC[["s1s0FC"]] ,
                              "stderr" = expC[["s0s1stderr"]]
        )
        
        TMPqpc <- data.frame( "Gene" = tmpG,
                              "Set" = "RT-qPCR",
                              "FC" = qCPRdataC[tmpG,][["relative.Werte.geoMean.C1C0.logFC"]] ,
                              #"stderr" = qCPRdataC[tmpG,][["relative.Werte.pooeld.var.C0C1_SatterthwaiteApproximation"]],
                              "stderr" = qCPRdataC[tmpG,][["s0s1stderr"]]
        )
        
        for (i in 1:nrow(TMPexp)) {
          tmp <- TMPqpc 
          tmp$ExpID <-  as.character(TMPexp[i,]$ExpID)
          TMP <- rbind(TMP , rbind(TMPexp[i, ],tmp[,  colnames(TMPexp) ]))
        }
      }  
      # TMP <- TMP[ TMP$Gene != 'KIF5C',]
      TMP$pid <- paste0(TMP$Gene,'::',TMP$ExpID)
      TMP$pid <- factor(TMP$pid)
      return(TMP)
      
    }   
    
    make.plot.data.exp.Ill.qPCR <- function(plotGens,MeanFoldChangeClass, class="c"){
      TMP <- data.frame()
      for (tmpG in qCPRdataC$rn  ) {
        expC <- MeanFoldChangeClass[[class]][qgenesIDs[[tmpG]], ]
        TMPexp1 <- data.frame("Gene" = tmpG,
                              "ExpID" = expC$rn,
                              'Set' = paste0(class,'1'),
                              "Mean" = expC$s1M ,
                              "stderr" = expC$s1stderr)
        TMPexp0 <- data.frame("Gene" = tmpG,
                              "ExpID" = expC$rn,
                              'Set' = paste0(class,'0'),
                              "Mean" = expC$s0M ,
                              "stderr" = expC$s0s1stderr)
        
        TMP <-  rbind(TMP,rbind(TMPexp0,TMPexp1) )
        
      }  
      # TMP <- TMP[ TMP$Gene != 'KIF5C',]
      TMP$pid <- paste0(TMP$Gene,'::',TMP$ExpID)
      TMP$pid <- factor(TMP$pid)
      return(TMP)
      
    }
    
    ### ggplot2 theme  ----------------------------------------------------------------------
      cols <- c(RColorBrewer::brewer.pal(9,"RdGy")[8],RColorBrewer::brewer.pal(9,"Reds")[6])
      base_size <- 12
      .thememap <- function(base_size = 12, legend_key_size = 0.4, base_family = "", col = "grey70") {
        ggplot2::theme_gray(base_size = base_size, base_family = base_family) %+replace% 
          ggplot2::theme(title = ggplot2::element_text(face="bold", colour=1,angle=0           ,vjust= 0.0,           size=base_size),
                         axis.title.x = ggplot2::element_text(face="bold", colour=1, angle=0   ,vjust= 0.0,           size=base_size),
                         # axis.text.x  = ggplot2::element_text(face="bold", colour=1, angle = -30 , vjust = 1, hjust = 0, size=base_size),
                         axis.text.x  = ggplot2::element_text(face="bold", colour=1, angle = 270          , hjust = 0, size=base_size),
                         strip.text.x = ggplot2::element_text(face="bold", colour=1, angle=0    ,vjust= 0.5,           size=base_size),
                         axis.title.y = ggplot2::element_text(face="bold", colour=1, angle=90   ,vjust= 1.5, hjust=.5, size=base_size),
                         axis.text.y  = ggplot2::element_text(face="bold", colour=1,                                  size=base_size),
                         legend.text  = ggplot2::element_text(face="bold" ,colour=1, angle=0  ,vjust= 0.0,             size=base_size),
                         legend.title = ggplot2::element_text(face="bold" ,colour=1, angle=0  ,vjust= 0.2,             size=base_size),
                         
                         
                         axis.ticks =  ggplot2::element_line(colour = "grey70", size = 0.5),
                         panel.grid.major =  ggplot2::element_line(colour = col, size = 0.2),
                         panel.grid.minor =  ggplot2::element_blank(),
                         panel.background = ggplot2::element_rect(fill="white",size = 0.2,),
                         # #panel.grid.minor.y = element_line(size=3),
                         # panel.grid.major = ggplot2::element_line(colour = "white"),
                         
                         # Force the plot into a square aspect ratio
                         # aspect.ratio = 1,
                         aspect.ratio = 9 / 16,
                         
                         legend.key.size  = ggplot2::unit(legend_key_size, "cm"),
                         plot.title = ggplot2::element_text(hjust = 0.5 , vjust= 1)          
          )
      }
```

## Barplot mean expression data of Illumina 
```{r barplot exp data ,echo=FALSE}       
    ### bar Illumina expression ----------------------------------------------------------------------
      PlotDataEXP <- make.plot.data.exp.Ill.qPCR(qCPRdata,MeanFoldChangeClass)
      
      # print(PlotDataEXP)
      
      rns <- levels(PlotDataEXP$Gene)[c(2,5,6,1,3,4)] ; PlotDataEXP$Gene <- factor(PlotDataEXP$Gene, levels = rns)
      rns <- levels(PlotDataEXP$pid)[c(2,5,6,1,3,4)]  ; PlotDataEXP$pid <- factor(PlotDataEXP$pid, levels = rns)
      
      cols <-  RColorBrewer::brewer.pal(8,'Paired')[c(6,2)] 
      limits <- aes(ymax = PlotDataEXP$Mean + PlotDataEXP$stderr , ymin=PlotDataEXP$Mean - PlotDataEXP$stderr)
      dodge <- position_dodge(width=0.9)
      
      PlotExp <- ggplot(PlotDataEXP,aes(x=factor(Gene),y=Mean,fill=factor(Set))) +  
        geom_bar(position="dodge",stat="identity") + 
        geom_errorbar(limits, position=dodge, width=0.25) +
        ylim(c(0 ,10)) +
        scale_fill_manual(values = cols,name="") + 
        labs(x = "", y = "Log2 mean expression",title="mean expression data of Illumina") +  
        .thememap(base_size = base_size,legend_key_size = 0.6) + 
        theme(legend.position="bottom") 
      print(PlotExp) 

```                
```{r results="PlotDataEXP", echo = FALSE}
knitr::kable(PlotDataEXP, caption = "Table 1", row.names = F)
```

## Barplot comparison of Illumina data and RT-qPCR Log2-fold change
```{r barplot FC, echo=FALSE}
    ### bar log2 FC qPCR Illumina  ----------------------------------------------------------------------
      PlotDataFC <- make.plot.data.FC.Ill.qPCR(qCPRdata,MeanFoldChangeClass)
      
      # print(PlotDataFC)
      
      rns <- levels(PlotDataFC$Gene)[c(2,5,6,1,3,4)] ; PlotDataFC$Gene <- factor(PlotDataFC$Gene, levels = rns)
      rns <- levels(PlotDataFC$pid)[c(2,5,6,1,3,4)]  ; PlotDataFC$pid <- factor(PlotDataFC$pid, levels = rns)
      
      cols <- RColorBrewer::brewer.pal(11,"PRGn")[c(2,10)]
      limits <- aes(ymax = PlotDataFC$FC + PlotDataFC$stderr , ymin=PlotDataFC$FC - PlotDataFC$stderr)
      dodge <- position_dodge(width=0.9)
      
      # g1 <- ggplot(PlotDataFC,aes(x=factor(pid),y=FC,fill=factor(Set))) +
      PlotFC <- ggplot(PlotDataFC,aes(x=factor(Gene),y=FC,fill=factor(Set))) +  
        geom_bar(position="dodge",stat="identity") + 
        geom_errorbar(limits, position=dodge, width=0.25) +
        ylim(c(-2 ,2)) +
        labs(x = "", y = "Log2-fold change",title="Comparison of Illumina data and RT-qPCR") +
        scale_fill_manual(values = cols,name="") + 
        .thememap(base_size = base_size,legend_key_size = 0.6) + 
        theme(legend.position="bottom") 
      print(PlotFC)     
```

```{r results="PlotDataFC", echo = FALSE}
knitr::kable(PlotDataFC, caption = "Table 2", row.names = F)
```


###  How expression pattern von group c should look like 
```{r group c example ,echo = FALSE }   
      ### bar log2 FC qPCR Illumina  ----------------------------------------------------------------------
        Leset <- get.Lset()
        pmat <- t(matrix(c(0,1,0,0,0,1),2,3))
        dimnames(pmat) <- list( c('no RNAi','RNAi by siRNA ALL','RNAi by siRNA I'),c('no EGF','EGF'))  
        pmat <- pmat[c(1,3,2),] ## to get same oder as in paper
        bk = seq(0,max(1),by = 1)
        hmcols <- c(RColorBrewer::brewer.pal(9,"Reds")[6],RColorBrewer::brewer.pal(9,"Blues")[7])# colorRampPalette(c("blue", "red"))(length(bk))
        
        print('A')
        pheatmap::pheatmap(pmat,color = hmcols,breaks = seq(-1,1,by = 1) ,display_numbers = T,fontsize_number = 20,number_format = '%.0f',cluster_rows = FALSE,cluster_cols = FALSE,main='expression pattern group c',fontsize = 12,gaps_row = c(1,2),gaps_col = 1,border_color = 'black',number_color = 'black',legend = FALSE)
```

###  How expression pattern of compare genes
```{r exp genes ,echo = FALSE }                  
        library(gridExtra)
        library(grid)
        library(gtable)
      
        Gene.gtable <- list()
        for( tmoG in names(qgenesIDs)){
          pmat <- t(matrix(normData$E[qgenesIDs[[tmoG]],],2,3))
          dimnames(pmat) <- list( c('no RNAi','RNAi by siRNA ALL','RNAi by siRNA I'),c('no EGF','EGF'))  
          pmat <- pmat[c(1,3,2),] ## to get same oder as in paper
          
          pmat <- 2^pmat
          tmpnScore <-  ceiling(pmat/100)*100
          # tmpnScore <-  ceiling(pmat/10)*10
          
          bk = seq(0,max(tmpnScore),by = 1)
          hmcols <- colorRampPalette(c("white", "darkred"))(length(bk)-1)
          #pheatmap::pheatmap(pmat,color = hmcols,breaks = seq(0,max(tmpnScore),by = 1) ,display_numbers = T,fontsize_number = 12,cluster_rows = FALSE,cluster_cols = FALSE,main=tmoG,fontsize = 12,gaps_row = c(1,2),gaps_col = 1,border_color = 'black',number_color = 'black', )
                             # filename = paste0(plotDir,'/Heatmap_Score_',tair,'.pdf') ,width = 10 ,height = 10)
          Gene.gtable[[tmoG]] <- pheatmap::pheatmap(pmat,silent = T,color = hmcols,breaks = seq(0,max(tmpnScore),by = 1) ,display_numbers = T,fontsize_number = 15,cluster_rows = FALSE,cluster_cols = FALSE,main=tmoG,fontsize = 12,gaps_row = c(1,2),gaps_col = 1,border_color = 'black',number_color = 'black')
        }
```

```{r exp genes big, echo=FALSE, fig.height=8, fig.width=12}
  addBorder_gtable <- function(g){ g <- gtable_add_grob(g,
                                            grobs = rectGrob(gp = gpar(fill = NA, lwd = 1)),
                                            t= 1 , b = nrow(g), l = 1, r = ncol(g))
                      # g <- gtable_add_grob(g,
                      #                      grobs = rectGrob(gp = gpar(fill = NA, lwd = 1)),
                      #                      t = 1, l = 1, r = ncol(g))
                      
                      return(g)
        }
  grid.arrange(
    addBorder_gtable(Gene.gtable$CKAP2L$gtable),
    addBorder_gtable(Gene.gtable$ROCK1$gtable),
    addBorder_gtable(Gene.gtable$TPR$gtable),
    addBorder_gtable(Gene.gtable$ALDH4A1$gtable),
    addBorder_gtable(Gene.gtable$CLCA2$gtable),
    addBorder_gtable(Gene.gtable$GALNS$gtable),
    nrow=2,ncol=3)             
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
