---
title: "Reproducible script for the publication"
author: "Weinholdt Claus"
date: "2018-05-29"
output: github_document 
---

<!--   html_document:
    toc: true
    theme: united
  pdf_document:
    toc: true
    highlight: zenburn -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Claus Weinholdt, Henri Wichmann, Johanna Kotrba, David H. Ardell, Matthias Kappler, Alexander W. Eckert, Dirk Vordermark and Ivo Grosse __Regulatory targets of epidermal growth factor receptor alternative isoforms in a glioblastoma cell line__

## Contents

* [Getting started](#start)
    * [Installation](#Installation)
    * [Data](#data)
* [Analysis](#Analysis)
    * [Normalizing Illumina BeadChips expression data](#Normalizing)
        * [Schematic Expression Pattern](#SchematicE)
    * [Calculating the log-likelihood for each gene in each group](#CalculatingLik)
    * [Calculating Bayesian Information Criterion of the log-likelihood ](#BIC)
    * [Approximating posterior by the Bayesian Information Criterion](#Posterior)    
* [Identification of genes belonging to group c](#Comapre)
    * [Examples for group c](#GrC)
    * [Fold changes plot of Illumina data vs RT-qPCR](#FCplot)

## <a name="start"></a>Getting started

### <a name="Installation"></a>Installation of necessary packages

Install and load packages containing the functions for the analyses

```r
install.packages("devtools")
devtools::install_github("GrosseLab/BGSC")
library(BGSC)
```

### <a name="data"></a> Load expression data
<!-- Loading data  -->
```r
data(ExpData)
```

```{r, comment=NA, echo=FALSE}
suppressWarnings( library("BGSC",quietly = TRUE,warn.conflicts = FALSE) )
library(ggplot2)
library(gridExtra)
library(grid)
library(gtable)
suppressWarnings( library(purrr,quietly = TRUE,warn.conflicts = FALSE) ) 
```

## <a name="Analysis"></a> Run analysis

### <a name="Normalizing"></a> Normalizing Illumina BeadChips expression data

We use the function *neqc* function from the *limma* package which was developed for normalizing Illumina BeadChips data.
The *neqc* function performs background correction using negative control probes followed by quantile normalization using negative and positive control probes. 
The *Illumina GenomeStudio* calculates and reports a detection p-value, which represents the confidence that a given transcript is expressed above background defined by negative control probes.
For further analysis, we used only those probes for which the detection p-values for all six probes was below 0.05.
```{r normalizeExpData}
normData <- normalizeExpData()
```
#### <a name="SchematicE"></a> Schematic Expression Pattern
We define that:

- Genes of the group $a$ are never regulated by EGF, whereas genes of groups b-d are regulated by EGF.
- Genes of the group $b$ are regulated by EGF only through other receptors besides EGFR isoforms.
- Genes of the group $c$ are regulated by EGFR isoforms II-IV and not by other receptors.
- Genes of the group $d$ are regulated by EGFR isoform I and not other receptors or EGFR isoforms II-IV.

<!--  Based on this reduction, we can now formulate the goal of this work as the identification of putative target genes regulated by EGFR isoforms II-IV and not by other receptors or more crisply as the goal of identifying genes of group $c$. -->

```{r schematic gr, echo=FALSE, fig.height=10, fig.width=10}
addBorder_gtable <- function(g){ g <- gtable_add_grob(g, grobs = rectGrob(gp = gpar(fill = NA, lwd = 1)),
                                                          t= 1 , b = nrow(g), l = 1, r = ncol(g))
      # g <- gtable_add_grob(g,
      #                      grobs = rectGrob(gp = gpar(fill = NA, lwd = 1)),
      #                      t = 1, l = 1, r = ncol(g))
      
      return(g)
    }
Lsets <- get.Lset()
IndicatorVar <- lapply(Lsets,function(x){
  vec <- rep(0,6);names(vec) <- c(1:6)
  if (!is.null(x$s1)) { vec[x$s1] <- 1 }
  return(vec)
})
IndicatorVar.gtable <- purrr::map2( IndicatorVar , names(IndicatorVar), function(vec,.y){
       pmat <- t(matrix(vec,2,3))
  dimnames(pmat) <- list( c('no RNAi','RNAi by siRNA ALL','RNAi by siRNA I'),c('no EGF','EGF'))  
  pmat <- pmat[c(1,3,2),] ## to get same oder as in paper
  bk = seq(0,max(1),by = 1)
  #hmcols <- c(RColorBrewer::brewer.pal(9,"Reds")[6],RColorBrewer::brewer.pal(9,"Blues")[7])# colorRampPalette(c("blue", "red"))(length(bk))
  hmcols <- RColorBrewer::brewer.pal(8,'Paired')[c(6,2)] 
  
  if(.y == 'a') hmcols <- 'slategray'
  
  pheatmap::pheatmap(pmat,silent = T,color = hmcols,breaks = seq(-1,1,by = 1) ,display_numbers = T,fontsize_number = 16,number_format = '%.0f',cluster_rows = FALSE,cluster_cols = FALSE,main=paste0('Expression pattern of for group ', .y),fontsize = 12 ,gaps_row = c(1,2),gaps_col = 1,border_color = 'black',number_color = 'black',legend = FALSE) 
} )

grid.arrange(
  addBorder_gtable(IndicatorVar.gtable$a$gtable),
  addBorder_gtable(IndicatorVar.gtable$b$gtable),
  addBorder_gtable(IndicatorVar.gtable$c$gtable),
  addBorder_gtable(IndicatorVar.gtable$d$gtable),
  nrow = 2,ncol = 2,
 top = textGrob("Expression pattern",gp = gpar(fontsize = 20,fontface = "bold")),
 bottom = textGrob("For group a all six expression values are assigned to the same class (black).\n For the groups b - d all six expression values are assigned to two classes - class 0 (red) and class 1 (blue).",gp = gpar(fontsize = 12,font = 1))
 ) 
   
``` 

### <a name="CalculatingLik"></a> Calculating the log-likelihood for each gene in each group (a, b, c, and d)
For group $a$ we assume that all six expression levels stem from the same Gaussian distribution. In this case ,the mean $\mu$ and standard deviation $\sigma$ of this Gaussian distribution (black) is equal to $\mu$ and $\sigma$ of the six expression levels. For the groups b-d, we assume that all six expression levels stems from a mixture of two Gaussian distributions with independent means $\mu_0$ and $\mu_1$, and one pooled standard deviation $\sigma$. For the groups b-d, we assume that the expression levels [$x_1$, $x_3$, and $x_5$], [$x_1$, $x_3$, $x_4$, and $x_5$], and [$x_1$, $x_3$, $x_4$, $x_5$, and $x_6$] stem from the Gaussian distribution based on $\mu_0$ (red), respectively. For the groups b-d, we assume that the expression levels [$x_2$, $x_4$, and $x_6$], [$x_2$ and $x_4$], and [$x_2$] stem from the Gaussian distribution based on $\mu_1$ (blue), respectively.

```{r logLik}
Lsets <- get.Lset()
normDataLogLikData <- logLikelihoodOfnormData(normData$E)
normDataLogLik <- normDataLogLikData[['logL']]
ALL.MUs  <- normDataLogLikData[['ALL.MUs']]
ALL.VARs <- normDataLogLikData[['ALL.VARs']]
```

#### <a name="ProbabilityDens"></a> Probability density plots of the Gaussian distributions
As an example, we show for each group a gene having the minimum log-likelihood. For the groups $a-d$, the examples are ABCB7, ACSL1, TPR, and ADAR, respectively. In each figure, we plot the probability density of the Gaussian distribution for the group $a$ as a black curve and mark the six log2-expression values with black circles. For groups $b−d$, we plot with red and blue curves the probability densities of the Gaussian distributions and mark the six log2-expression values with circles, which are colored according to classes for class 0 in red and for class 1 in blue.
```{r Density, echo=TRUE, fig.height=9, fig.width=9}
GeneExample <- c('ILMN_1687840','ILMN_1684585','ILMN_1730999','ILMN_2320964') 
names(GeneExample) <- c('a','b','c','d')
tmpPlot <- purrr::map2(GeneExample,names(GeneExample),function(.x,.y) Density.NV.fit.plot(id = .x ,normData, ALL.MUs, ALL.VARs, useGroup = .y ,DOplot = FALSE) )
grid.arrange( tmpPlot$a + theme(legend.position = "none"),
              tmpPlot$b + theme(legend.position = "none"),
              tmpPlot$c + theme(legend.position = "none"),
              tmpPlot$d + theme(legend.position = "none") ,ncol=2,nrow=2)
```

### <a name="BIC"></a> Calculating Bayesian Information Criterion of the log-likelihood 
Performing classification through model selection based on minimum log-likelihood is problematic when the number of free model parameters is not identical among all models under comparison. Here, model $a$ has two free model parameters, while models $b$, $c$, and $d$ have three. Hence, a naive classification based on a minimum log-likelihood criterion would give a spurious advantage to models $b$, $c$, and $d$ with three free model parameters over model $a$ with only two free parameters.To eliminate that spurious advantage, we compute marginal likelihoods $p(x|z)$ using the approximation of Schwarz et al. commonly referred to as Bayesian Information Criterion.

```{r BIC}
npar <- sapply(Lsets, function(x) sum(!sapply(x,is.null ) )) + 1  ## number parameters for log-likelihood -> mean + var 
k <-  sapply(Lsets, function(x) sum( sapply(x,length) )) ## number of samples
print(rbind(npar,k))
normDataBIC <- get.IC(normDataLogLik , npar, k , IC = 'BIC')
``` 

```{r results="PlotDataEXP", echo = FALSE}
print('Number of genes assigned to group with the minimal Bayesian Information Criterion')
BICminInd <- apply( normDataBIC,MARGIN = 1, FUN = minIndex)
tmp <- table(BICminInd); names(tmp) <- names(Lsets)
tmpMat <- as.matrix(tmp);colnames(tmpMat) <- "#genes assigned to group"
print(t(tmpMat))
```


### <a name="Posterior"></a> Approximating posterior by the Bayesian Information Criterion
We assume that $70\%$ of all genes are not regulated by EGF, so we define the prior probability for group a by $p(a) = 0.70$. Further, we assume that the remaining $30\%$ of the genes fall equally in groups with EGF-regulation, so we define the prior probabilities for groups $b$, $c$, and $d$ by $p(b) = p(c) = p(d) = 0.1$. 
We can compute for $z \in \{a, b, c, d\}$ the posterior $p(z|x) \approx p(x|z) \cdot p(z)$ and then perform Bayesian model selection by assigning each gene to that group $z$ with the maximum approximate posterior $p(z|x)$.
```{r Posterior}
normDataPosterior <- get.Posterior( normDataBIC ,Pis = c(0.7,0.1,0.1,0.1))
POSTmaxInd <- apply(normDataPosterior, MARGIN = 1 ,FUN = maxIndex)
PostClass <- get.gene.group(data = normDataPosterior,indexing = "maximal",filter = 0.75, DoPlot = TRUE)
```   

## <a name="Comapre"></a> Identification of genes belonging to group c
Genes of the group $c$ are putative target genes regulated by EGFR isoforms II-IV and not by other receptors.

### <a name="GrC"></a> Examples for group c
After calculating the log2-fold change for group $c$ by $\mu_{c1}$ - $\mu_{c0}$, we validated three up-regulated genes, namely CKAP2L, ROCK1, and TPR and three down-regulated genes, namely ALDH4A1, CLCA2, and GALNS.

```{r load_qPCR}
    ### calculating mean and log2-fold change
    MeanFoldChangeClass <- get.log2Mean.and.log2FC(normData = normData) 
    
    ### load qPCR data
    qCPRdataC <- getQPCR()
    
    ### annotation of gene examples
    IDs.dt <- data.table::data.table(normData$genes,keep.rownames = T,key = 'rn')
    IDs.dt.c <- IDs.dt[rownames(PostClass$resFilter$c),]
    data.table::setkey(IDs.dt.c,'SYMBOL')
    
    qgenesIDs <- lapply(qCPRdataC$rn, function(qg) as.character(IDs.dt.c[qg,][['rn']]) )
    names(qgenesIDs) <- qCPRdataC$rn
```

<!-- 
### Schematic expression patterns of group c 
c0 is red and c1 is blue ... 
```{r group c example, echo=FALSE, fig.height=4, fig.width=5}
      ### bar log2 FC qPCR Illumina  ----------------------------------------------------------------------
        Leset <- get.Lset()
        pmat <- t(matrix(c(0,1,0,0,0,1),2,3))
        dimnames(pmat) <- list( c('no RNAi','RNAi by siRNA ALL','RNAi by siRNA I'),c('no EGF','EGF'))  
        pmat <- pmat[c(1,3,2),] ## to get same oder as in paper
        bk = seq(0,max(1),by = 1)
        #hmcols <- c(RColorBrewer::brewer.pal(9,"Reds")[6],RColorBrewer::brewer.pal(9,"Blues")[7])# colorRampPalette(c("blue", "red"))(length(bk))
        hmcols <- RColorBrewer::brewer.pal(8,'Paired')[c(6,2)] 
        pheatmap::pheatmap(pmat,color = hmcols,breaks = seq(-1,1,by = 1) ,display_numbers = T,fontsize_number = 20,number_format = '%.0f',cluster_rows = FALSE,cluster_cols = FALSE,main='expression pattern group c',fontsize = 10,gaps_row = c(1,2),gaps_col = 1,border_color = 'black',number_color = 'black',legend = FALSE)
```
-->

####  Expression patterns 
```{r GroupC_ExpGenes ,echo = FALSE }                  

        Gene.gtable <- list()
        for(tmoG in names(qgenesIDs)){
          pmat <- t(matrix(normData$E[qgenesIDs[[tmoG]],],2,3))
          dimnames(pmat) <- list( c('no RNAi','RNAi by siRNA ALL','RNAi by siRNA I'),c('no EGF','EGF'))  
          pmat <- pmat[c(1,3,2),] ## to get same oder as in paper
          
          pmat <- 2^pmat
          tmpnScore <-  ceiling(pmat/100)*100
          # tmpnScore <-  ceiling(pmat/10)*10
          
          bk = seq(0,max(tmpnScore),by = 1)
          hmcols <- colorRampPalette(c("white", "darkred"))(length(bk)-1)
          #pheatmap::pheatmap(pmat,color = hmcols,breaks = seq(0,max(tmpnScore),by = 1) ,display_numbers = T,fontsize_number = 12,cluster_rows = FALSE,cluster_cols = FALSE,main=tmoG,fontsize = 12,gaps_row = c(1,2),gaps_col = 1,border_color = 'black',number_color = 'black', )
                             # filename = paste0(plotDir,'/Heatmap_Score_',tair,'.pdf') ,width = 10 ,height = 10)
          Gene.gtable[[tmoG]] <- pheatmap::pheatmap(pmat,silent = T,color = hmcols,breaks = seq(0,max(tmpnScore),by = 1) ,display_numbers = T,fontsize_number = 15,cluster_rows = FALSE,cluster_cols = FALSE,main=tmoG,fontsize = 12,gaps_row = c(1,2),gaps_col = 1,border_color = 'black',number_color = 'black')
        }
```

```{r groupC_ExpGenesPlot, echo=FALSE, fig.height=8, fig.width=12}
  addBorder_gtable <- function(g){ g <- gtable_add_grob(g,
                                            grobs = rectGrob(gp = gpar(fill = NA, lwd = 1)),
                                            t= 1 , b = nrow(g), l = 1, r = ncol(g))
                      # g <- gtable_add_grob(g,
                      #                      grobs = rectGrob(gp = gpar(fill = NA, lwd = 1)),
                      #                      t = 1, l = 1, r = ncol(g))
                      
                      return(g)
        }
  grid.arrange(
    addBorder_gtable(Gene.gtable$CKAP2L$gtable),
    addBorder_gtable(Gene.gtable$ROCK1$gtable),
    addBorder_gtable(Gene.gtable$TPR$gtable),
    addBorder_gtable(Gene.gtable$ALDH4A1$gtable),
    addBorder_gtable(Gene.gtable$CLCA2$gtable),
    addBorder_gtable(Gene.gtable$GALNS$gtable),
    nrow=2,ncol=3)             
```

#### Barplot of Illumina expression data

```{r MeanIlluminaExpression, echo=FALSE}
    tmp <- data.table(MeanFoldChangeClass$c[do.call(c,qgenesIDs),.(rn,s1M,s0M,s1s0FC)])
    setnames(tmp,c('rn','s1M','s0M','s1s0FC'),c('IlluminaID',"meanC1","meanC0",'log2-fold change'))
    tmp$GeneName <- names(qgenesIDs)
    tmp <- tmp[,c('GeneName','IlluminaID',"meanC1","meanC0",'log2-fold change'),with=F]
    setkey(tmp,'GeneName')
    print(tmp[c('CKAP2L', 'ROCK1','TPR','ALDH4A1', 'CLCA2', 'GALNS'),])
```

```{r barplotExpData ,echo=FALSE}       
    ### bar Illumina expression ----------------------------------------------------------------------
      PlotDataEXP <- make.plot.data.exp.Ill.qPCR(qCPRdata = qCPRdataC, MeanFoldChangeClass = MeanFoldChangeClass,class="c")
      
      # print(PlotDataEXP)
      
      rns <- levels(PlotDataEXP$Gene)[c(2,5,6,1,3,4)] ; PlotDataEXP$Gene <- factor(PlotDataEXP$Gene, levels = rns)
      rns <- levels(PlotDataEXP$pid)[c(2,5,6,1,3,4)]  ; PlotDataEXP$pid <- factor(PlotDataEXP$pid, levels = rns)
      
      cols <-  RColorBrewer::brewer.pal(8,'Paired')[c(6,2)] 
      limits <- aes(ymax = PlotDataEXP$Mean + PlotDataEXP$stderr , ymin=PlotDataEXP$Mean - PlotDataEXP$stderr)
      dodge <- position_dodge(width = 0.9)
      
      base_size <- 12
      PlotExp <- ggplot(PlotDataEXP,aes(x=factor(Gene),y=Mean,fill=factor(Set))) +  
        geom_bar(position="dodge",stat="identity") + 
        geom_errorbar(limits, position=dodge, width=0.25) +
        ylim(c(0 ,10)) +
        scale_fill_manual(values = cols,name="") + 
        labs(x = "", y = "Log2 mean expression",title="mean expression data of Illumina") +  
        thememapBarplot(base_size = base_size,legend_key_size = 0.6) + 
        # theme( aspect.ratio = 9 / 16) + 
        theme(legend.position="bottom") 
      print(PlotExp) 

```

<!-- 
```{r results="PlotDataEXP", echo = FALSE}
knitr::kable(PlotDataEXP, caption = "Table 1", row.names = F)
```
-->

### <a name="FCplot"></a> Log2 fold changes of Illumina data vs RT-qPCR
We have found that the six $\log_2$-fold changes of the Illumina microarray expression levels, and those of the qPCR expression levels show a Pearson correlation coefficient of $0.99$ (p-value = 0.00002). Therefor, we can suggest that the set of 1,140 genes might contain some further putative target genes of isoforms II-IV of the epidermal growth factor receptor in tumor cells.

```{r barplotFC, echo=FALSE}
    ### bar log2 FC qPCR Illumina  ----------------------------------------------------------------------
      PlotDataFC <- make.plot.data.FC.Ill.qPCR(qCPRdata = qCPRdataC, qgenesIDs = qgenesIDs, MeanFoldChangeClass, class="c" )
      
      print(cor.test( PlotDataFC[PlotDataFC$Set == 'Microarray','FC'],PlotDataFC[PlotDataFC$Set == 'qPCR','FC'] ))    
  
      # print(PlotDataFC)
      
      rns <- levels(PlotDataFC$Gene)[c(2,5,6,1,3,4)] ; PlotDataFC$Gene <- factor(PlotDataFC$Gene, levels = rns)
      rns <- levels(PlotDataFC$pid)[c(2,5,6,1,3,4)]  ; PlotDataFC$pid <- factor(PlotDataFC$pid, levels = rns)
      
      cols <- RColorBrewer::brewer.pal(11,"PRGn")[c(2,10)]
      limits <- aes(ymax = PlotDataFC$FC + PlotDataFC$stderr , ymin=PlotDataFC$FC - PlotDataFC$stderr)
      dodge <- position_dodge(width=0.9)
      
      # g1 <- ggplot(PlotDataFC,aes(x=factor(pid),y=FC,fill=factor(Set))) +
      PlotFC <- ggplot(PlotDataFC,aes(x=factor(Gene),y=FC,fill=factor(Set))) +  
        geom_bar(position="dodge",stat="identity") + 
        geom_errorbar(limits, position=dodge, width=0.25) +
        ylim(c(-2 ,2)) +
        labs(x = "", y = "Log2-fold change",title="Comparison of Illumina data and RT-qPCR") +
        scale_fill_manual(values = cols,name="") + 
        thememapBarplot(base_size = base_size,legend_key_size = 0.6) + 
        # theme( aspect.ratio = 9 / 16) + 
        theme(legend.position="bottom") 
      print(PlotFC)     
```

```{r results="PlotDataFC", echo = FALSE}
PlotDataFC.dt <- data.table(PlotDataFC)
setkeyv(PlotDataFC.dt,c('Gene','Set'))
knitr::kable(PlotDataFC.dt[ c('CKAP2L', 'ROCK1','TPR','ALDH4A1', 'CLCA2', 'GALNS'),.(Gene,Set,FC,stderr)], caption = "Table 1", row.names = F,digits = 2)
```

```{r sessionInfo, echo=FALSE}
print( sessionInfo() )
```

<!--  Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot. -->
